---
title: "tidycensus_tinkering"
author: "Colleen Minnihan"
date: "4/14/2021"
output: html_document
---

#load packages
```{r}
library(tidycensus)
library(tidyverse)
library(sf)
```

#get API Key:  http://api.census.gov/data/key_signup.html.

Make sure you don't push with your key in plain sight!
```{r, install = TRUE}
#census_api_key("key")
```

```{r}
#in geography you can put us, region, state, county, school district (elementary, secondary, unified)

#to see variables
#View(load_variables(year = 2019, "acs5", cache = TRUE))
```


```{r}
#median income in last 12 months by county
get_acs(geography = "county",
        variables = "B06011_001",
        state = "MN",
        year = 2019)

#median income in last 12 months by school district (unified: K-12)
get_acs(geography = "school district (unified)",
        variables = "B06011_001",
        state = "MN",
        year = 2019)
```

```{r}
#visualizing median income in last 12 months by county
income_county <- get_acs(geography = "county",
        variables = "B06011_001",
        state = "MN",
        year = 2019)

income_county[1:25,] %>%
  ggplot(aes(x = estimate, y = reorder(NAME, estimate))) + 
  geom_point()
```


```{r}
#"Black or African American alone" race by school district
get_acs(geography = "school district (unified)",
        variables = "B02001_003",
        state = "MN",
        year = 2019)

#"White alone" race by school district
get_acs(geography = "school district (unified)",
        variables = "B02001_002",
        state = "MN",
        year = 2019)

#maybe we should convert these to percentages rather than a number (when we do data wrangling)
```

Visualize school districts in MN by median income
```{r}
district_income <- get_acs(state = "MN", geography = "school district (unified)", 
                  variables = "B06011_001", geometry = TRUE, year = 2019)

district_income %>%
  ggplot(aes(fill = estimate)) + 
  geom_sf(color = NA)
```

Visualize counties in MN by median income
```{r}
county_income <- get_acs(state = "MN", geography = "county", 
                  variables = "B06011_001", geometry = TRUE, year = 2019)

county_income %>%
  ggplot(aes(fill = estimate)) + 
  geom_sf(color = NA)
```

Potential variables to look at for the 2019 census:

**POPULATION**
  *total population (B01003_001)
  (perhaps the bigger districts are more likely to offer it)
  
**SEX**
  * Male 5-9 years old (B01001_004)
  * Male 10-14 years old (B01001_005)
  * Male 15-17 years old (B01001_006)
  (I would imagine that schools with a larger number of men may be more likely to have comp sci offerings?)
  
**RACE**
  * Race: white alone (B02001_002)
  * Race: Black or African American alone (B02001_003)
  * Race: Asian alone (B02001_005)
  * Race: Native Hawaiian and other pacific islander alone (B02001_006)
  * Race: Two or more races (B02001_008)
  
**INCOME/POVERTY**
  * Median household income in last 12 months (B19013_001)
  * receive supplemental security income, cash public assistance income, or food stamps in households w/ child under 18 (B09010_001)
  * number of earners in family (B19122_001)
  * median gross rent (B25031_001)
  * housing units w/ a mortgage (B25027_001)

**HOME ENVIRONMENT**
  * Prescence of computer and internet subscription in household (B28003_001)


#load ACT data
```{r}
act <- read_csv("act_data.csv")

#data cleaning
act_clean <-
act %>% 
  filter(`Grad Year` == 2018,
         `Analysis Level` == "District",
         `District Name` != "MINNESOTA DEPT OF EDUCATION",
         `Avg Comp` != ".") %>% 
  select(`District Name`, `Avg Comp`) %>% 
  mutate(`Dist Num` = as.integer(str_extract_all(`District Name`, "[:digit:]+"))) %>% 
  mutate(`Dist Num`= replace(`Dist Num`, `District Name`=="GRANADA-HUNTLEY-EAST CHAIN SD", 2536),
         `Dist Num`= replace(`Dist Num`, `District Name`=="ALBERT LEA AREA SCHOOLS", 241),
         `Dist Num`= replace(`Dist Num`, `District Name`=="ATWATER-COSMOS-GROVE CITY SD", 2396),
         `Dist Num`= replace(`Dist Num`, `District Name`=="BOLD SCHOOL DISTRICT", 2534),
         `Dist Num`= replace(`Dist Num`, `District Name`=="BUFFALO HANOVER MONTROSE SD", 0877),
         `Dist Num`= replace(`Dist Num`, `District Name`=="BUFFALO LAKE-HECTOR-STEWART SD", 2159),
         `Dist Num`= replace(`Dist Num`, `District Name`=="EASTERN CARVER CO SCHOOLS", 112),
         `Dist Num`= replace(`Dist Num`, `District Name`=="MANKATO AREA PUBLIC SCHS", 77),
         `Dist Num`= replace(`Dist Num`, `District Name`=="MINNEAPOLIS PUBLIC SCH DIST", 9991),
         `Dist Num`= replace(`Dist Num`, `District Name`=="MINNEOTA PUBLIC SCHOOLS", 414),
         `Dist Num`= replace(`Dist Num`, `District Name`=="PLAINVIEW ELGIN MILLVILLE SD", 2899),
         `Dist Num`= replace(`Dist Num`, `District Name`=="ROYALTON PUBLIC SCHOOLS", 485),
         `Dist Num`= replace(`Dist Num`, `District Name`=="WASECA PUBLIC SCHOOL DIST", 829),
         `Dist Num`= replace(`Dist Num`, `District Name`=="WAUBUN-OGEMA-WHITE EARTH PSD", 435),
         `Dist Num`= replace(`Dist Num`, `District Name`=="WINONA AREA PUBLIC SCHOOL DIST", 861)) 

act_clean
```

#Making a dataset with relevant census variables
```{r}

mn_2019_census <- get_acs(state = "MN", 
                         geography = "school district (unified)",
                         variables = c("B01003_001",
                                       "B01001_004",
                                       "B01001_005",
                                       "B01001_006",
                                       "B02001_002",
                                       "B02001_003",
                                       "B02001_005",
                                       "B02001_006",
                                       "B02001_008",
                                       "B19013_001",
                                       "B09010_001", 
                                       "B25031_001",
                                       "B25027_001",
                                       "B28003_001"),
                         geometry = TRUE,
                         year = 2019)

mn_2019_census_nogeom <- mn_2019_census %>%
  st_drop_geometry() %>% #get rid of the geometry so we can spread()
  select(-`moe`) %>% #get rid of moe variable(not sure what it is)
  spread(variable, estimate) %>% #give each variable a column
  rename(total_pop = B01003_001,
         male_5to9 = B01001_004,
         male_10to14 = B01001_005,
         male_15to17 = B01001_006,
         race_white_only = B02001_002,
         race_Black_only = B02001_003,
         race_Asian_only = B02001_005,
         race_pacific_islander_only = B02001_006,
         at_least_two_races = B02001_008,
         med_household_inc_12mo = B19013_001,
         SSI_pubassist_foodstamps = B09010_001,
         med_gross_rent = B25031_001,
         house_units_w_mortgage = B25027_001,
         comp_internet_subscrip_in_house = B28003_001
         ) %>% #rename variables
  mutate(District = str_extract(`NAME`, "[^,]+")) %>% #delete ", Minnesota" after district name
  select(-NAME)#remove `NAME` column
  

mn_2019_census_nogeom

```

#load dataset with district numbers
```{r}
districts <- read_csv("districts.csv")

districts <- districts %>%
  mutate(District = `District Name`) %>%
  select(-`X3`, - `District Name`)
```
#join so that census dataset has district numbers
```{r}
mn_2019_census_nogeom <- mn_2019_census_nogeom %>%
  left_join(districts)

mn_2019_census_nogeom

#the only NA value is for the "remainder of Minnesota"- should we get rid of this row anyways because it's not a school district?
colSums(is.na(mn_2019_census_nogeom))
```




